---
title: "Title for Group"
author: "Ehsan Habibpour, Natalie Short, and Ryan Altman"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


# Load necessary libraries

```{r}
library(tidycensus)
library(ggplot2)
library(dplyr)
library(tidyr)
library(writexl)
library(readr)
library(stringr)
```


## R Code 	
```{r}
year_tg <- 2023

# Set your Census API key
census_api_key("3504e2e4186afe0b674a6f3f54caefd9ef3af8d9", install = TRUE, overwrite=TRUE)
#census_api_key("34777786635d1f4f42107118005896f09cde55e1", install = TRUE, overwrite=TRUE)

## Use `load_variables()` function to explore more variables if needed
codelists <- load_variables(year_tg, "acs5", cache = TRUE)

#################################################################################
variables <- c(
  HIC_total = "B25141_001E",
  HIC_with_mortgage = "B25141_002E",
  HIC_with_mortgage_50 = "B25141_003E",
  HIC_with_mortgage_200 = "B25141_004E",
  HIC_with_mortgage_400 = "B25141_005E",
  HIC_with_mortgage_650 = "B25141_006E",
  HIC_with_mortgage_900 = "B25141_007E",
  HIC_with_mortgage_1250 = "B25141_008E",
  HIC_with_mortgage_1750 = "B25141_009E",
  HIC_with_mortgage_2250 = "B25141_010E",
  HIC_with_mortgage_2750 = "B25141_011E",
  HIC_with_mortgage_3250 = "B25141_012E",
  HIC_with_mortgage_3750 = "B25141_013E",
  HIC_with_mortgage_m4000 = "B25141_014E",
  HIC_without_mortgage = "B25141_015E",
  HIC_without_mortgage_50 = "B25141_016E",
  HIC_without_mortgage_200 = "B25141_017E",
  HIC_without_mortgage_400 = "B25141_018E",
  HIC_without_mortgage_650 = "B25141_019E",
  HIC_without_mortgage_900 = "B25141_020E",
  HIC_without_mortgage_1250 = "B25141_021E",
  HIC_without_mortgage_1750 = "B25141_022E",
  HIC_without_mortgage_2250 = "B25141_023E",
  HIC_without_mortgage_2750 = "B25141_024E",
  HIC_without_mortgage_3250 = "B25141_025E",
  HIC_without_mortgage_3750 = "B25141_026E",
  HIC_without_mortgage_m4000 = "B25141_027E",
  MOR_total = "B25096_001E",
  MOR_with = "B25096_002E",
  MOR_with_25000 = "B25096_003E",
  MOR_with_75000 = "B25096_004E",
  MOR_with_125000 = "B25096_005E",
  MOR_with_175000 = "B25096_006E",
  MOR_with_250000 = "B25096_007E",
  MOR_with_400000 = "B25096_008E",
  MOR_with_625000 = "B25096_009E",
  MOR_with_875000 = "B25096_010E",
  MOR_with_m1000000 = "B25096_011E",
  MOR_without = "B25096_012E",
  MOR_without_25000 = "B25096_013E",
  MOR_without_75000 = "B25096_014E",
  MOR_without_125000 = "B25096_015E",
  MOR_without_175000 = "B25096_016E",
  MOR_without_250000 = "B25096_017E",
  MOR_without_400000 = "B25096_018E",
  MOR_without_625000 = "B25096_019E",
  MOR_without_875000 = "B25096_020E",
  MOR_without_m1000000 = "B25096_021E"
)

census_data <- get_acs(
  geography = "county",
  variables = variables,
  year = year_tg,
  survey = "acs5", # Use 5-year ACS data
  output = "wide"
)

################################
# Extract the variable names
variable_columns <- names(variables)

# Keep only the specified variables and the 'NAME' column for county names
census_data <- census_data %>%
  select(NAME, all_of(variable_columns))
```

```{r}
# Extracting States and creating a new variable
census_data$state <- sub(".*,\\s*", "", census_data$NAME) # Gets everything after the last comma
census_data$county <- sub(",\\s*[^,]*$", "", census_data$NAME) # Gets everything before the last comma

# Then remove "County" if it exists at the end
census_data$county <- sub(" County$", "", census_data$county)
census_data$county <- sub(" Census Area$", "", census_data$county)
census_data$county <- sub(" Municipio$", "", census_data$county)
census_data$county <- sub(" Parish$", "", census_data$county)
census_data$county <- sub(" Borough$", "", census_data$county)
census_data$county <- sub(" Planning Region$", "", census_data$county)
census_data$county <- sub(" city$", "", census_data$county)
```

```{r}
# Move state and county back to the front

census_data <- census_data |>
  select(county, state, everything()) |>
  select(-"NAME")
```

NOAA storm and events data download:

```{r}
noaa_data <- read_csv("Data/county_events.csv")
noaa_data <- noaa_data |>
  filter(year == 2018 | year == 2023)
```

# Joining the census and NOAA datasets:

```{r}
# Adjusting county names for a join

census_data <- census_data |>
  mutate(county = str_to_upper(county, locale = "en")) |>
  group_by(state) |>
  arrange(county, .by_group = TRUE) |>
  ungroup()

# Pivoting the NOAA data so each county only appears once

noaa_1_county <- noaa_data |>
  pivot_wider(names_from = event_type, values_from = event_count, values_fill = 0,
              id_cols = c("county", "state", "year")) |>
  rename(state_bad = state) |>
  mutate(total_events = rowSums(across(where(is.numeric))) - year) |>
  group_by(state_bad) |>
  arrange(county, .by_group = TRUE) |>
  ungroup() |>
  mutate(county_state = paste(county, state_bad, sep = " -- ")) |>
  select(county, state_bad, county_state, year, total_events, everything())

head(census_data)
head(noaa_1_county)
```

### Splitting the NOAA event data by year -- not necessarily what we use, but it's good to have them split up.

```{r}
noaa_2018 <- noaa_1_county |>
  filter(year == 2018)

noaa_2023 <- noaa_1_county |>
  filter(year == 2023)

head(noaa_2018)
head(noaa_2023)
```

### Counting the number of counties in each 

```{r}
print("2018:")
nrow(noaa_2018)
print("2023:")
nrow(noaa_2023)
```

### Finding counties that only appear in one year.

```{r}

exclusive_counties_18 <- setdiff(noaa_2018$county_state, noaa_2023$county_state)
exclusive_counties_23 <- setdiff(noaa_2023$county_state, noaa_2018$county_state)
exclusive_counties <- c(exclusive_counties_18, exclusive_counties_23)
length(exclusive_counties)
```

```{r}

noaa_2018_bothyearcounties <- noaa_2018 |>
  filter(!(county_state %in% exclusive_counties))
  
noaa_2023_bothyearcounties <- noaa_2023 |>
    filter(!(county_state %in% exclusive_counties))

print("Checking the row count:")
print(nrow(noaa_2018_bothyearcounties))
print(nrow(noaa_2023_bothyearcounties))
```

# Exploratory Data Analysis

Are storms correlated by county?

```{r}
ggplot(mapping = aes(
  x = noaa_2018_bothyearcounties$total_events,
  y = noaa_2023_bothyearcounties$total_events,
  alpha = 0.5
)) +
  geom_point(color = 'goldenrod4', shape = "*", size = 5) +
  labs(
    title = "Five-Year Change in Weather Events",
    subtitle = "Counties Where Events > 0, per NOAA") +
  xlab("2018") +
  ylab("2023") +
  theme_bw() +
  theme(legend.position = "none") +
  theme(title = element_text(color = 'navy')) +
  coord_equal(ratio = 1, expand = TRUE, xlim = c(0, 750), ylim = c(0, 750))
```

Let's zoom in on the majority of counties:

```{r}
ggplot(mapping = aes(
  x = noaa_2018_bothyearcounties$total_events,
  y = noaa_2023_bothyearcounties$total_events,
  alpha = 0.5
)) +
  geom_point(color = 'goldenrod4', shape = "*", size = 5) +
  labs(
    title = "Five-Year Change in Weather Events",
    subtitle = "Counties Where Events > 0, per NOAA") +
  xlab("2018") +
  ylab("2023") +
  theme_bw() +
  theme(legend.position = "none") +
  theme(title = element_text(color = 'navy')) +
  coord_equal(ratio = 1, expand = TRUE, xlim = c(0, 310), ylim = c(0, 310))
```

…and zoom in even closer:

```{r}
ggplot(mapping = aes(
  x = noaa_2018_bothyearcounties$total_events,
  y = noaa_2023_bothyearcounties$total_events,
  alpha = 0.5
)) +
  geom_point(color = 'goldenrod4', shape = "*", size = 5) +
  labs(
    title = "Five-Year Change in Weather Events",
    subtitle = "Counties Where Events > 0, per NOAA") +
  xlab("2018") +
  ylab("2023") +
  theme_bw() +
  theme(legend.position = "none") +
  theme(title = element_text(color = 'navy')) +
  coord_equal(ratio = 1, expand = TRUE, xlim = c(0, 150), ylim = c(0, 150))
```

Delineating by event type:

```{r}
event_sum_bothyears <- noaa_1_county |>
  select(-c(county, state_bad, county_state)) |>
  pivot_longer(cols = everything()) |>
  group_by(name) |>
  summarise(instances = sum(value)) |>
  filter(name != "year", name != "total_events") |>
  rename(event_type = name)

event_sum_2018 <- noaa_2018 |>
  select(-c(county, state_bad, county_state)) |>
  pivot_longer(cols = everything()) |>
  group_by(name) |>
  summarise(instances = sum(value)) |>
  filter(name != "year") |>
  rename(event_type = name)

event_sum_2023 <- noaa_2023 |>
  select(-c(county, state_bad, county_state)) |>
  pivot_longer(cols = everything()) |>
  group_by(name) |>
  summarise(instances = sum(value)) |>
  filter(name != "year") |>
  rename(event_type = name)
```

Floods:

```{r}

floods <- c("flood", "flash flood", "coastal flood", "lakeshore flood")

floods_bothyears <- event_sum_bothyears |>
  filter(event_type %in% floods) |>
  arrange(desc(instances))

floods_2018 <- event_sum_2018 |>
  filter(event_type %in% floods) |>
  arrange(desc(instances))

floods_2023 <- event_sum_2023 |>
  filter(event_type %in% floods) |>
  arrange(desc(instances))

summary(floods_bothyears)

head(floods_bothyears)
```

