---
title: "Title for Group"
author: "Ehsan Habibpour, Natalie Short, and Ryan Altman"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Load necessary libraries

```{r}
library(tidyverse)
library(tidycensus)
library(dplyr)
library(tidyr)
library(writexl)
library(readr)
library(stringr)
library(plm)
library(ggplot2)
```

```{r}
# get census data
get_census_year_data <- function(year_tg, price_index, nt = 2) {
  # Define variables
  variables <- c(
    houseowners_total = "B25087_001E", #Housing units
    houseowners_with_mortgage = "B25087_002E", #Housing units with a mortgage
    houseowners_without_mortgage = "B25087_020E", #Housing units without a mortgage
    household_size_average = "B25010_001E", #Average Household Size
    income_median = "B19013_001E", #Median household income in the past 12 months
    house_value_median = "B25077_001E", #Median Value
    owner_cost_median = "B25088_001E", #Median selected monthly owner costs
    owner_cost_median_with_mortgage = "B25088_002E", #same with mortgage
    owner_cost_median_without_mortgage = "B25088_003E", #same without mortgage
    owner_cost_to_income_median = "B25092_001E", #sama as a percentage of yearly income
    owner_cost_to_income_median_with_mortgage = "B25092_002E", #same with mortgage
    owner_cost_to_income_median_without_mortgage = "B25092_003E" #same without mortgage
  )

  census_data <- get_acs(
    geography = "county",
    variables = variables,
    year = year_tg,
    survey = "acs5",
    output = "wide"
  )

  # Process data
  census_data <- census_data %>%
    select(GEOID, NAME, all_of(names(variables))) %>%
    mutate(
      state_fips = as.numeric(substr(GEOID, 1, 2)),    # Extract first 2 digits for state FIPS
      county_fips = as.numeric(substr(GEOID, 3, 5)),   # Extract last 3 digits for county FIPS
      state = sub(".*,\\s*", "", NAME),
      county = sub(",\\s*[^,]*$", "", NAME)
    ) %>%
    mutate(
      county = str_remove(county, " County$"),
      county = str_remove(county, " Census Area$"),
      county = str_remove(county, " Municipio$"),
      county = str_remove(county, " Parish$"),
      county = str_remove(county, " Borough$"),
      county = str_remove(county, " Planning Region$"),
      county = str_remove(county, " city$")
    ) %>%
    mutate(
      income_median = income_median/price_index,
      house_value_median = house_value_median/price_index,
      owner_cost_median = owner_cost_median * 12/price_index,
      owner_cost_median_with_mortgage = owner_cost_median_with_mortgage * 12/price_index,
      owner_cost_median_without_mortgage = owner_cost_median_without_mortgage * 12/price_index,
      year = year_tg
    ) %>%
    select(
      GEOID, state_fips, county_fips, state, county, year, houseowners_total,
      houseowners_with_mortgage, houseowners_without_mortgage, household_size_average,
      house_value_median, owner_cost_median, income_median,
      owner_cost_median_with_mortgage, owner_cost_median_without_mortgage,
      owner_cost_to_income_median, owner_cost_to_income_median_with_mortgage,
      owner_cost_to_income_median_without_mortgage
    )
  
  return(census_data)
}

# Get data for each year
census_data_2023 <- get_census_year_data(2023, 1) 
census_data_2018 <- get_census_year_data(2018, 0.824)
census_data_2013 <- get_census_year_data(2013, 0.765)

# Combine all years into one dataset
census_data_total <- bind_rows(
  census_data_2023,
  census_data_2018,
  census_data_2013
)

# Sort the final dataset
census_data_total <- census_data_total %>%
  arrange(state, county, year)
```

NOAA storm and events data download:

```{r}
noaa_data <- read_csv("Data/county_events.csv")
noaa_data <- noaa_data |>
  filter(year == 2013 | year == 2018 | year == 2023)
```

# Joining the census and NOAA datasets:

```{r}
# Adjusting county names for a join

census_data_total <- census_data_total |>
  mutate(county = str_to_upper(county, locale = "en")) |>
  group_by(state) |>
  arrange(county, .by_group = TRUE) |>
  ungroup()

# Pivoting the NOAA data so each county only appears once

noaa_1_county <- noaa_data |>
  pivot_wider(names_from = event_type, values_from = event_count, 
              values_fill = list(event_count = 0),
              id_cols = c("state_fips_code", "cz_fips_code", "year", "state", "county"))

```

```{r}

noaa_1_county <- noaa_1_county |>
  rename(state_bad = state) |>
  mutate(total_events = rowSums(across(where(is.numeric))) - year) |>
  group_by(state_fips_code) |>
  arrange(county, .by_group = TRUE) |>
  ungroup() |>
  mutate(county_state = paste(county, state_fips_code, sep = " -- ")) |>
  mutate(fips_ovr = paste(county, as.character(state_fips_code), as.character(cz_fips_code), sep = "__")) |>
  select(county, fips_ovr, total_events, state_fips_code, county_state, year, everything())

head(census_data_total)
head(noaa_1_county)
```

### Splitting the NOAA event data by year -- not necessarily what we use, but it's good to have them split up.

```{r}
noaa_2013 <- noaa_1_county |>
  filter(year == 2013)

noaa_2018 <- noaa_1_county |>
  filter(year == 2018)

noaa_2023 <- noaa_1_county |>
  filter(year == 2023)

head(noaa_2013)
head(noaa_2018)
head(noaa_2023)
```

### Counting the number of counties in each 

```{r}
print("2013:")
nrow(noaa_2013)
print("2018:")
nrow(noaa_2018)
print("2023:")
nrow(noaa_2023)
```

### Finding counties that only appear in one year.

```{r}

exclusive_counties_13 <- setdiff(noaa_2013$fips_ovr, noaa_2018$fips_ovr)
exclusive_counties_13_23 <- setdiff(noaa_2013$fips_ovr, noaa_2023$fips_ovr)
exclusive_counties_13_18_1 <- setdiff(noaa_2023$fips_ovr, noaa_2013$fips_ovr)
exclusive_counties_13_23_1 <- setdiff(noaa_2018$fips_ovr, noaa_2013$fips_ovr)
exclusive_counties_18 <- setdiff(noaa_2018$fips_ovr, noaa_2023$fips_ovr)
exclusive_counties_23 <- setdiff(noaa_2023$fips_ovr, noaa_2018$fips_ovr)
exclusive_counties <- c(exclusive_counties_13, exclusive_counties_13_23, 
                        exclusive_counties_18, exclusive_counties_23,
                        exclusive_counties_13_18_1, exclusive_counties_13_23_1)
length(exclusive_counties)
```

```{r}
noaa_2013_bothyearcounties <- noaa_2013 |>
  filter(!(fips_ovr %in% exclusive_counties))

noaa_2018_bothyearcounties <- noaa_2018 |>
  filter(!(fips_ovr %in% exclusive_counties))
  
noaa_2023_bothyearcounties <- noaa_2023 |>
    filter(!(fips_ovr %in% exclusive_counties))

print("Checking the row count:")
print(nrow(noaa_2013_bothyearcounties))
print(nrow(noaa_2018_bothyearcounties))
print(nrow(noaa_2023_bothyearcounties))
```

# Exploratory Data Analysis

Are storms correlated by county?

```{r}
ggplot(mapping = aes(
  x = noaa_2018_bothyearcounties$total_events,
  y = noaa_2023_bothyearcounties$total_events,
  alpha = 0.5
)) +
  geom_point(color = 'goldenrod4', shape = "*", size = 5) +
  labs(
    title = "Five-Year Change in Weather Events",
    subtitle = "Counties Where Events > 0, per NOAA") +
  xlab("2018 | Total Events") +
  ylab("2023 | Total Events") +
  theme_bw() +
  theme(legend.position = "none") +
  theme(title = element_text(color = 'navy')) +
  coord_equal(ratio = 1, expand = TRUE, xlim = c(0, 750), ylim = c(0, 750))
```

Let's zoom in on the majority of counties:

```{r}
ggplot(mapping = aes(
  x = noaa_2018_bothyearcounties$total_events,
  y = noaa_2023_bothyearcounties$total_events,
  alpha = 0.5
)) +
  geom_point(color = 'goldenrod4', shape = "*", size = 5) +
  labs(
    title = "Five-Year Change in Weather Events",
    subtitle = "Counties Where Events > 0, per NOAA") +
  xlab("2018 | Total Events") +
  ylab("2023 | Total Events") +
  theme_bw() +
  theme(legend.position = "none") +
  theme(title = element_text(color = 'navy')) +
  coord_equal(ratio = 1, expand = TRUE, xlim = c(0, 310), ylim = c(0, 310))
```

…and zoom in even closer:

```{r}
ggplot(mapping = aes(
  x = noaa_2018_bothyearcounties$total_events,
  y = noaa_2023_bothyearcounties$total_events,
  alpha = 0.5
)) +
  geom_point(color = 'goldenrod4', shape = "*", size = 5) +
  labs(
    title = "Five-Year Change in Weather Events",
    subtitle = "Counties Where Events > 0, per NOAA") +
  xlab("2018 | Total Events") +
  ylab("2023 | Total Events") +
  theme_bw() +
  theme(legend.position = "none") +
  theme(title = element_text(color = 'navy')) +
  coord_equal(ratio = 1, expand = TRUE, xlim = c(0, 150), ylim = c(0, 150))
```

A linear model for this relationship:

```{r}
model <- lm(noaa_2023_bothyearcounties$total_events ~ noaa_2018_bothyearcounties$total_events)
summary(model)
```
2018 events and 2023 events are highly correlated by county, which makes sense because weather is generally stable within a 5-year period.  Overall, though, there is a very slight positive relationship between year-change and total events -- demonstrating climate change and the greater prevalence of extreme weather events.

Delineating by event type:

```{r}
event_sum_bothyears <- noaa_1_county |>
  select(-c(county, state_bad, county_state, fips_ovr)) |>
  pivot_longer(cols = everything()) |>
  group_by(name) |>
  summarise(instances = sum(value)) |>
  filter(name != "year", name != "total_events", name != "state_fips_code", name != "cz_fips_code") |>
  rename(event_type = name)

event_sum_2018 <- noaa_2018 |>
  select(-c(county, state_bad, county_state, fips_ovr)) |>
  pivot_longer(cols = everything()) |>
  group_by(name) |>
  summarise(instances = sum(value)) |>
  filter(name != "year", name != "state_fips_code", name != "cz_fips_code") |>
  rename(event_type = name)

event_sum_2023 <- noaa_2023 |>
  select(-c(county, state_bad, county_state, fips_ovr)) |>
  pivot_longer(cols = everything()) |>
  group_by(name) |>
  summarise(instances = sum(value)) |>
  filter(name != "year", name != "state_fips_code", name != "cz_fips_code") |>
  rename(event_type = name)
```

Floods:

```{r}

floods <- c("flood", "flash flood", "coastal flood", "lakeshore flood")

floods_bothyears <- event_sum_bothyears |>
  filter(event_type %in% floods) |>
  arrange(desc(instances))

floods_2018 <- event_sum_2018 |>
  filter(event_type %in% floods) |>
  arrange(desc(instances))

floods_2023 <- event_sum_2023 |>
  filter(event_type %in% floods) |>
  arrange(desc(instances))

summary(floods_bothyears)

head(floods_bothyears)
```

```{r}

census_data_total <- census_data_total |>
  mutate(county = str_to_upper(county, locale = "en")) |>
  mutate(fips_ovr = paste(county, as.character(state_fips), as.character(county_fips), sep = "__"))

census2013 <- census_data_total |>
  filter(year == 2013)

census2018 <- census_data_total |>
  filter(year == 2018)

census2023 <- census_data_total |>
  filter(year == 2023)

exclusive_counties_census <- setdiff(census2023$county, noaa_2023_bothyearcounties$county)
exclusive_counties_noaa <- setdiff(noaa_2023_bothyearcounties$county, census2023$county)
exclusive_counties <- c(exclusive_counties_census, exclusive_counties_noaa)
length(exclusive_counties)

census2023_noaa2023 <-
  full_join(noaa_2023_bothyearcounties, census2023, 
            by = "fips_ovr")

census2023_noaa2023 <- census2023_noaa2023 |>
  filter(!(county.x %in% exclusive_counties)) |>
  filter(!(county.y %in% exclusive_counties))
```

```{r}
main_set <-
  full_join(noaa_1_county, census_data_total,
            by = join_by(year == year, fips_ovr == fips_ovr))

main_set <- main_set |>
  filter(!(county.x %in% exclusive_counties)) |>
  filter(!(county.y %in% exclusive_counties)) |>
  select(fips_ovr, state, year, total_events, owner_cost_median, everything()) |>
  drop_na()
```

# Models and Analyses

```{r}
model_1 <- lm(owner_cost_median ~ total_events, data = main_set)
summary(model_1)
```

A graphical representation of the above:

```{r}
ggplot(data = main_set, aes(
  x = total_events,
  y = owner_cost_median,
  color = cut(year, breaks = c(2012, 2014, 2017, 2019, 2022, 2024)),
  alpha = 0.5
)) +
  geom_point() +
  theme_bw() +
  labs(title = "Major Weather Events' Effect on Costs") +
  labs(subtitle = "By County, By Year") + 
  xlab("Total Major Weather Events in Year") +
  ylab("Owner Cost Median ($)") +
  scale_color_manual(values = c('green', 'darkred', 'blue'), 
                     labels = c("2013", "2018", "2023")) +
  theme(legend.title = element_blank()) +
  theme(plot.title = element_text(color = "navy")) +
  theme(plot.subtitle = element_text(color = "navy")) +
  guides(alpha = "none")
```
Create lag 
```{r}
library(dplyr)
library(plm)

main_set <- main_set %>%
  mutate(ending_fips_ovr = sub("^[^_]*_","",fips_ovr))%>%
  select(ending_fips_ovr, everything())


panel_main_set <- main_set %>%
  arrange(fips_ovr, year) %>%   
  group_by(fips_ovr) %>%      
  mutate(
    lag_events = lag(total_events, n = 1) )%>%
    select(lag_events, year, total_events, fips_ovr, everything())%>%
    ungroup()

panel_main_set <- panel_main_set %>%
    mutate(log_owner_cost_median=log(owner_cost_median))

pooled_ols_model2 <- plm(log_owner_cost_median ~lag_events, 
                  data = panel_main_set,
                  model="pooling")
